BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2024-07-09 manuscript/manuscript_APPS.rmd
focusing intro

Nearly 400 million specimens are housed in herbaria around the world (@thiers2021herbaria).
These specimens, collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity, have found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
However, The rate of accessioning new collections to herbaria diminished in the 20th century as priorities in biology shifted away from describing, documenting, and understanding earths diversity towards understanding cellular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion). 
Fortunately, renewed interest in collections have brought herbaria of all sizes back to the forefront of plant sciences (@ronsted2020integrative, @marsico2020small).

In fact recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
These current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative). 
Further we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, and electronic sensing (@tosa2021rapid). 
Even within systematics, the ability to collect and analyze micro-morphological data has fostered revisionary studies in many groups providing synapomorphies which can be correlated to molecular results to improve the interpretation of monophyletic clades.

However, while there is recognized need for more specimens the skills of collecting and processing specimens, and time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial). 
The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience.  
While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection and preparing specimens for herbaria. 
This scenario results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for some specimens to ever be accessioned at all. 
Problems which young collectors face generally include both the lack of dedicated time awarded to them at a seasons end to process specimens, and a general lack of formal education on cartography, natural history, taxonomy, and plant systematics.

The successful generation of an herbarium specimen includes many steps which are easy to take for granted. 
For example, while the acquisition of political information for a collection site appears simple, it is only so if the collector has the adequate resources at their disposal. 
Given the association of boundaries with topographically complex areas (e.g. watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g. ArcMap, which is relatively expensive at 100$ year), or more likely Google Maps by individual site a time consuming process. 
This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.

@@

time_directions_grabber <- system.time({ # write directions to sites
  SoS_gkey = Sys.getenv("Sos_gkey")
  data1 <- directions_grabber(data, api_key = SoS_gkey)
})

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied. BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` record in error.

@@

\begin{wrapfigure}{r}{0.45\textwidth}
  \centering
    \includegraphics[width=0.45\textwidth]{../graphics/tables/Screenshot-DataSources.png}
  \caption{Sources of Data required for operations}
\end{wrapfigure}

\newpage
# CONCLUSIONS
BarnebyLives is a tool which is able to rapidly acquire relevant geographic, and taxonomic data. It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data. The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.

>>>>

Nearly 400 million specimens are housed in herbaria around the globe (@thiers2021herbaria).
These specimens, collected to describe the taxonomic diversity of plants and document the worlds floristic diversity, have recently found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity towards understanding cellular and molecular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices (@funk2014erosion). 
Fortunately, renewed interest approaches in collections generated by 'big data approaches' have brought herbarium collections back to the forefront of the natural sciences (@ronsted2020integrative, @marsico2020small).

In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative). 
Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid). 
While image or purely observational (rather than collections based) citizen science initializes (e.g. iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images. 
Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes, and seeds or pollen; forever ensuing specimens as the ultimate data source to center most efforts around.

However, while there is recognized need for more specimens the skills of collecting and processing specimens and the time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial). 
The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience.  
While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection and preparing specimens for herbaria. 
This scenario results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for some specimens to ever be accessioned at all. 
Problems which young collectors face generally include both the lack of dedicated time awarded to them at a seasons end to process specimens, and a general lack of formal education on cartography, natural history, taxonomy, and plant systematics.

The successful generation of an herbarium specimen includes many steps which are easy to take for granted. 
For example, while the acquisition of political information for a collection site appears simple, it is only so if the collector has the adequate resources at their disposal.

Given the association of boundaries with topographically complex areas (e.g. watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g. ArcMap, which is relatively expensive at $100 year), or more likely Google Maps by individual site a time consuming process. 
This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.

@@

time_directions_grabber <- system.time({ # write directions to sites
  SoS_gkey = Sys.getenv("Sos_gkey")
  data <- directions_grabber(data, api_key = SoS_gkey)
})

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied. BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.

@@

\begin{wrapfigure}{r}{0.45\textwidth}
  \centering
    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
  \caption{Sources of Data required for operations}
\end{wrapfigure}

# CONCLUSIONS
BarnebyLives is a tool which is able to rapidly acquire relevant geographic, and taxonomic data. It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data. The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
